\chapter
 [NetQASM: A low-level instruction set architecture for hybrid quantum-classical programs in a quantum internet]
 {NetQASM: A low-level instruction set architecture for hybrid quantum-classical programs in a quantum internet}
\label{chp:netqasm}

\acrodef{CNPU}{Classical Network Processing Unit}
\acrodef{QNPU}{Quantum Network Processing Unit}
\acrodef{NetQASM}{Network Quantum Assembly Language}
\acrodef{CQC}{Classical Quantum Combiner}

\newcommand{\IMMEDIATE}{\textbf{IMMEDIATE}}
\newcommand{\REGISTER}{\textbf{REGISTER}}
\newcommand{\ADDRESS}{\textbf{ADDRESS}}
\newcommand{\ARRAYENTRY}{\textbf{ARRAY\_ENTRY}}
\newcommand{\ARRAYSLICE}{\textbf{ARRAY\_SLICE}}


\begin{abstract}
    We introduce NetQASM, a low-level instruction set architecture for quantum
    internet applications. NetQASM is a universal, platform-independent and
    extendable instruction set with support for local quantum gates, powerful
    classical logic and quantum networking operations for remote entanglement
    generation. Furthermore, NetQASM allows for close integration of classical
    logic and communication at the application layer with quantum operations at
    the physical layer. This enables quantum network applications to be
    programmed in high-level platform-independent software, which is not
    possible using any other QASM variants. We implement NetQASM in a series of
    tools to write, parse, encode and run NetQASM code, which are available
    online. Our tools include a higher-level SDK in Python, which allows an easy
    way of programming applications for a quantum internet. Our SDK can be used
    at home by making use of our existing quantum simulators, NetSquid and
    SimulaQron, and will also provide a public interface to hardware released on
    a future iteration of Quantum Network Explorer.
\end{abstract}

\blfootnote{
This chapter is extracted from the article:
\fullcite{dahlberg_2022_netqasm}.

Contributions: TODO
}


\input{chapters/main/netqasm/introduction}
\input{chapters/main/netqasm/model}
\input{chapters/main/netqasm/use-cases}
\input{chapters/main/netqasm/design-considerations}
\input{chapters/main/netqasm/design-decisions}
\input{chapters/main/netqasm/implementation}
\input{chapters/main/netqasm/python_sdk}
\input{chapters/main/netqasm/evaluation}
\input{chapters/main/netqasm/conclusion}

\section{Data availability}
The data that support the findings of this study are openly available at the following DOI:
\url{https://doi.org/10.4121/21355329}.
The software packages created for this work (and used for the simulation) are
\url{https://github.com/QuTech-Delft/netqasm} (NetQASM) and \url{https://github.com/QuTech-Delft/squidasm} (SquidASM).

\begin{xstretch}
\printbibliography[heading=subbibintoc,title={References},notcategory=noprint]
\end{xstretch}
