\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}
\setheader{Acknowledgments}

This thesis is the result of more than four years of research and development, and would not have been possible without the help of many people.
During my PhD at QuTech, I have had the privilege of being surrounded by a diverse group of smart, interesting, and hard-working people.
All people who I've interacted with have shaped, either directly or more indirectly, the way I've worked through, sometimes battled through, my PhD.

First of all, many thanks to my promotor \textbf{Stephanie Wehner}.
It was an honor to have been a part of working towards your vision for a quantum internet.
I admire your ability to keep up many, many balls (although I'm not jealous of your busy calendar)
and I enjoyed the balance you gave me between freedom and responsibility as well as giving very valuable feedback and guidance when needed.
Although I've at times felt more an engineer than a scientist, you have always been supportive and motivated me about continuing the PhD, even when I had serious doubts.
I look forward to see where your ambition for the quantum internet leads to, and I'm glad I'll remain a part of the wider endeavour.

I would also like to thank my committee members for taking the time to read through my thesis and subject me to critical questions during my defense:
\textbf{Arie van Deursen}, acting as second promotor;
\textbf{Rodney van Meter}, joining online all the way from Japan;
\textbf{Michele Amoretti}, who I've had the pleasure of interacting with in various QIA- and compiler-discussion-related settings;
\textbf{Tim Coopmans}, who was a fellow PhD student at QuTech when I joined and I have the honor to have him in this new role;
\textbf{Rob Kooij}, who specifically moved his calendar around so he could join the defense;
and \textbf{Koen Langendoen}, for being available as a reserve member.

Thanks to \textbf{Soham Chakraborty} and \textbf{Johannes Borregaard} for being in my Go/No Go commitee.

Of course many thanks to my wonderful paranymphs, \textbf{Carlo} and \textbf{Francisco}.
Carlo, it has been a pleasure sharing an office for a long time. You really have a genuine interest in other people.
Francisco, I admire your sharp wits! Looking forward to all the use cases you'll cook up for the quantum internet.

Special thanks to \textbf{Axel} who, in the chaotic situation of the initial lockdown and a supervisor on maternity leave, managed to on-board me and inspire me in my first months of the PhD.
Your work has been a great starting point for my own PhD.
% The first two years of my PhD were mostly during (partial) lockdowns and other restrictions and therefore mostly remote.

Other people in my group that I've had the pleasure of interacting with during the beginning of my PhD include
\textbf{Aram}, \textbf{Matt}, \textbf{Guus} and \textbf{David}.

\textbf{\'Alvaro}, \textbf{Bethany}, \textbf{Scarlett}, you're all nearing the end of your PhD, and you can all be proud of what you've achieved!
\textbf{F\'elix}, \textbf{Janice}, \textbf{Jeroen}, \textbf{Kaushik}, \textbf{Luca}, \textbf{Margrete}, \textbf{Soubhadra}, \textbf{Sounak}, \textbf{Thomas}, \textbf{Tzula}, thanks for being awesome group members, and I'm expecting great things from all of you.
\textbf{Diego}, with your enthousiasm for code development you have been a great addition to our group.
\textbf{Sam} and \textbf{Sacha}: I'm looking forward to the cool things you're going to do with Qoala!

Not only my own group but also other people at QuTech have inspired me.
\textbf{Eric}, \textbf{Fenglei}, \textbf{Hemant}, \textbf{Kenneth}, \textbf{S\'ebastian}, \textbf{Siddanth}, thanks for being either great office neighbours, good chess opponents, or just nice colleagues!

Special thanks to \textbf{Mariagrazia}, I really enjoyed doing the QNodeOS experiments together and we can be very proud about the result!
Also thanks to \textbf{Przemek}, \textbf{Tracy} and all other co-authors of the QNodeOS paper.

I also feel very privileged for all the trips abroad that I was allowed to make, for conferences and research visits.
From Bad Honnef to Barcelona, from Austin to Paris, and from Bali to Seattle, those trips always left me inspired and with renewed energy.
Special thanks to \textbf{Carmina} and \textbf{Anabel} for hosting me in Valencia.
\textbf{Aritra}, \textbf{Gayane}, \textbf{Luise}, \textbf{Matt}, \textbf{Medina}, \textbf{Pablo}, and \textbf{Sebastian}, I had a lot of fun with you in Seattle.

Thanks to the master students and interns that I supervised - you all helped me in my own work.
\textbf{Bob}, your project really helped kickstart the Qoala project.
% Emir?
\textbf{Atak}, it was a pleasure working with you and you contributed greatly to the Qoala simulator.
\textbf{Hana}, thanks for being such a great master student that I actually didn't have to do much myself!
I'm looking forward to working more with you in the Stack Team.

Speaking about the Stack Team,
\textbf{Guilherme}, \textbf{Ingmar}, \textbf{Thom} and \textbf{Wojciech}, you all have been crucial in the development of QNodeOS, and I'm glad I can keep working with some of you in the future!

Thanks also to all other software engineers at QuTech that have contributed to projects I've been involved in.
\textbf{Michal}, thanks for taking over the SquidASM development! It's in very capable hands.
Thanks to \textbf{Animesh}, \textbf{Bobby}, \textbf{Fer}, \textbf{Giuliano}, \textbf{Ivo}, \textbf{Mark}, \textbf{Olaf}, \textbf{Ravi V} and other software engineers that have helped with QNE.
\textbf{Paul} and \textbf{Ravi B}, we haven't worked on any projects together, but I enjoyed hanging out with you in Leuven.

A big shout-out to the \textbf{QuTech band (T2 stars, formerly Q2)}.
Thanks for the great rehearsals every Wednesday and all the awesome performances that we did.
Over the years there have been too many members to list them all here, but one thing is for sure: you're all insanely talented and I immensely enjoy playing with you.

This PhD would also not have been possible without great ways to spend time outside of QuTech.
Special thanks to my fellow \textbf{Tiramisu} band members, \textbf{David}, \textbf{Floris} and \textbf{Pim}.
I can always really unwind by making music and having fun together with you.
We'll make that album one day!
\textbf{Grace}, also with you I really enjoyed playing piano together---especially in the lockdown period this was a great way to distract myself from the PhD.
Shout-out to the \textbf{DSBA-USSR badminton club} for keeping me fit and helping me meet new friends.
Thanks also to \textbf{Arminius} for the inspiring events as well as cultural trips to Romania and Suriname.
I'd also like to mention \textbf{Boris}, \textbf{Dirk}, \textbf{Jonas}, \textbf{Luke}, \textbf{Marianne}, \textbf{Mark} and \textbf{Shuham} for intellectual (but also just very fun) philosophical discussions in our book club.

\textbf{Elke}, \textbf{Jeroen}, \textbf{Rob}, \textbf{Sjoerd} en \textbf{Tom}, ik ben blij dat we nog steeds onze vriendengroep hebben sinds de middelbare school!
Uiteraard noem ik ook mijn familie, \textbf{Papa}, \textbf{Mama} en \textbf{Peter}, op wie ik altijd kan terugvallen.
Ook vond ik altijd steun bij \textbf{Jan}, \textbf{Annet}, \textbf{Opa}, \textbf{Oma}, \textbf{Lense}, \textbf{Machteld}, \textbf{Hugo}, en \textbf{Martijn}.
En als laatste natuurlijk mijn lieve \textbf{Anne}, mijn grote steun, met jouw liefde kan ik alles aan.



% Gustavo, Sander, Luc, Timo, Laurens, Hanifa, Maia, Desa, Pablo, "Q2 band" (or should I say already T2* ?)

% Tiramisu
% Boekenclub
% Arminius
% Badminton


% \begin{flushright}
% {\itshape
% Bart \\
% Delft, Maart 2025.
% }
% \end{flushright}
