\begin{center}

{\Large\titlefont\bfseries Propositions}

\medskip

accompanying the dissertation

\medskip

% Print the title.
{\makeatletter
\titlestyle\bfseries\large\@title
\makeatother}

% Print the optional subtitle.
{\makeatletter
\ifx\@subtitle\undefined\else
\titlefont\titleshape\@subtitle
\fi
\makeatother}

\medskip

by

\medskip

% Print the full name of the author.
\makeatletter
{\large\titlefont\bfseries\@firstname\ {\titleshape\@lastname}}
\makeatother

\end{center}

\bigskip

\begin{enumerate}[widest=10]
  % \item The quantum internet is a solution looking for problems.

  % \item The name of a tool or algorithm is more important than the thing it does.

  % \item There should be no more page limits on papers since LLMs can summarize them for you.

  % \item An applied scientist is just an engineer using big words.
  % % vague line between engineering and science

  % \item Quantum computing development speed will be limited severely as long as one does not provide a solution to the problem whether a qubit refers to a value or a memory location.

  % \item Science should remain in balance with engineering otherwise it risks losing touch with what is really needed.

  % \item Having an opinion about everything either shows an overconfidence or a waste of energy.

  % Wisdom as opposed to just knowledge/better tech
  % Working hours, 4 day work week, burnout, stress
  % Iets met religie/spiritualiteit
  % neoliberalisme: everything is a competition and market; you even have to sell/promote yourself 
  % 4-year PhD is not science; it's an engineering project due to limited time in which you have to forcefully produce "output" (no time to do proper science)
  % finiteness of life, limitations
  % latex shouldn't be used anymore, it's ancient

  % STRONG (still to be re-formulated)
  % thesis-related
  \item Existing full-stack models for classical computing and networking seduce scientists into trying to apply similar models to quantum computing and networking prematurely.
  \item For increasing quantum network application performance (such as success probability), improving scheduling is far more effective than improving compilation.
  
  % not thesis-related
  \item Program code supporting scientific work should undergo the same scrutiny as the manuscript text when evaluating publication.
  \item Researchers should be encouraged to improve their papers using large language models (LLMs) such as ChatGPT.
  \item Whether a result `advances science' depends not on some objective standard, but rather the scientist's skill in convincing the reader that their work is non-trivial.
  \item The proposal that science may at some point provide a `theory of everything' is naive at best and harmful for humankind at worst.
  \item When proposing a new paradigm, scientific evaluation in the form of `it works' is sufficient.
  \item A Doctor of Philosophy (PhD) should have a thorough grasp of philosophy in order to be worthy of their title.
  \item Treating life as a project severely lowers its quality.
  \item Traffic lights are configured in such a way that at an intersection, at any point in time, there is at least one vehicle waiting at a red light.

\end{enumerate}

\bigskip \noindent
Propositions 1 and 2 pertain to this thesis.

% The following text is dictated by the promotieregelement.
\bigskip \noindent
These propositions are regarded as opposable and defendable, and have been approved as such by the
promotors prof.\ dr.\ S.\ D.\ C.\ Wehner and prof.\ dr.\ A.\ van Deursen.
