\chapter*{List of Publications}
\addcontentsline{toc}{chapter}{List of Publications}
\setheader{List of Publications}

\def\aicon{\makebox[1.25em][l]{\footnotesize \faUser}}
\def\ficon{\makebox[1.25em][l]{\footnotesize \faFileTextO}}
\def\picon{\makebox[1.25em][l]{\footnotesize \faPencilSquareO}}

\begin{enumerate}[label={\arabic*.},itemsep=0.7\baselineskip,parsep=0.3\baselineskip]
    \item[6.] \textbf{B. van der Vecht*}, S. Oslovich*, D. Rivera, and S. Wehner. \textit{Qoala Compiler}. In preparation.
        \\ \ficon This article is included in this thesis as \cref{chp:compiler}.

    \item[5.]
        \textbf{B. van der Vecht*}, A. T. Yücel, H. Jirovská, and S. Wehner.
        \textit{Qoala: an Application Execution Environment for Quantum Internet Nodes}.
        arXiv preprint:2502.17296.
        {\small \textsc{doi:}} \texttt{10.48550/arXiv.2502.17296}.
        \\ \ficon This article is included in this thesis as \cref{chp:qoala}.

    \item[4.]
        C. D. Donne*, M. Iuliano*, \textbf{B. van der Vecht*}, G. M. Ferreira, H. Jirovská, T. van der Steenhoven, A. Dahlberg, M. Skrzypczyk, D. Fioretto, M. Teller, P. Filippov, A. R.-P. Montblanch, J. Fischer, B. van Ommen, N. Demetriou, D. Leichtle, L. Music, H. Ollivier, I. t. Raa, W. Kozlowski, T. Taminiau, P. Pawełczak, T. Northup, R. Hanson, and S. Wehner.
        \textit{Design and demonstration of an operating system for executing applications on quantum network nodes}.
        % July 25, 2024.
        arXiv preprint:2407.18306.
        {\small \textsc{doi:}} \texttt{10.48550/arXiv.2407.18306}.
        To appear in Nature,
        {\small \textsc{doi:}} \texttt{10.1038/s41586-025-08704-w}.
        \\ \ficon This article is included in this thesis as \cref{chp:qnodeos}.

    \item[3.]
        M. Pompili*, C. Delle Donne*, I. te Raa, \textbf{B. van der Vecht}, M. Skrzypczyk, G. M. Ferreira, L. de Kluijver, A. J. Stolk, S. L. N. Hermans, P. Pawełczak, W. Kozlowski, R. Hanson, and S. Wehner.
        \textit{Experimental Demonstration of Entanglement Delivery Using a Quantum Network Stack}.
        In: Nature Partner Journal (npj) Quantum Information 8.1 (2022), p. 121.
        {\small \textsc{doi:}} \texttt{10.1038/s41534-022-00631-2}.

    \item[2.]
        QuTech.
        \textit{Quantum Network Explorer Software Development Kit (QNE SDK)}.
        Non-scientific publication.
        \url{https://www.quantum-network.com}.
        2021.

    \item[1.]
        A. Dahlberg*, \textbf{B. van der Vecht*}, C. Delle Donne, M. Skrzypczyk, I. te Raa, W. Kozlowski, and S. Wehner.
        \textit{NetQASM---A Low-Level Instruction Set Architecture for Hybrid Quantum-Classical Programs in a Quantum Internet}.
        In: Quantum Science and Technology 7.3 (2022), p. 035023.
        {\small \textsc{doi:}} \texttt{10.1088/2058-9565/ac753f}.
        \\ \ficon This article is included in this thesis as \cref{chp:netqasm}.
\end{enumerate}

\vskip 5mm
\centerline{*Main author}

\begin{xstretch}
\printbibliography[heading=subbibintoc,title={References},notcategory=noprint]
\end{xstretch}
